The Multi-Stencil Language

### 2016-01-22 ARTICLES/IJPP/article/parallelism.tex
article IJPP add reduction Helene 22/01/2016

\begin{mydef}
A synchronization between two computations $k_i \pprec k_j$ is defined as a specific computation 
\begin{equation*}
sync(S,R,w,comp)
\end{equation*}
\end{mydef}
where $S=\emptyset$, $R=\{(w_i,n_j)\}$, $w=(w_i,n_j)$ and $comp=identity$. In other words, $w_i$ has to be synchronized for the neighborhood $n_j$.

\begin{mydef}
If $k_i \pprec k_j$, $k_j$ is replaced by the list
\begin{equation*}
[sync(\emptyset,\{(w_i,n_j)\},(w_i,n_j),identity), k_j]
\end{equation*}
\end{mydef}

@@

\begin{mydef}
For a reduction kernel $k(S,R,(w,D),comp)$, if we denote by $w^i$, $0 \leq i<p$, the local $w$ computed on the resource $i$, a reduction synchronization is defined as the specific computation 
\begin{equation*}
sync(S,R_s,w,comp)
\end{equation*}
\end{mydef}
where, $S=\emptyset$, $R=\{(w^0,entity(w^0)) \dots (w^{p-1},entity(w^{p-1}))\}$ and $w$ and $comp$ are the same than the actual reduction.

@@

\begin{mydef}
From the ordered list of computation $\Gamma$, a new ordered list $\Gamma_{sync}$ is obtained from the call $\Gamma_{sync} = F_{sync}(\Gamma,0)$, where $F_{sync}$ is the recursive function defined as
\begin{equation*}
F_{sync}(\Gamma,j) = 
\begin{cases} 	[] \mbox{ if }j=|\Gamma|\\
				[sync(\emptyset,\{w_i\},w_i,identity,\emptyset), k_j] \cdot F_{sync}(\Gamma,j+1)\\
				\text{ } \qquad \forall i<j \text{, and }k_i \pprec k_j\\
				[k_j, sync(\emptyset,,w_j,comp_j)] \cdot F_{sync}(\Gamma,j+1)\\ 
				\text{ } \qquad \mbox{if } w_j \in \mathcal{S} \text{, and } R_j=\{(r_j,entity(r_j))\}\\
				[\Gamma[j]] \cdot F_{sync}(\Gamma,j+1) \text{ else}.
\end{cases}
\end{equation*}
\end{mydef}

% The construction of $\Gamma_{data}$ implies that if a synchronization is needed it is always proceeded before the computation wich needs it. One can notice that a combination of synchronizations is possible in some cases, and thus it is possible to synchronize more than one data at a time. %If the number of synchronizations is reduced, final performance of the program will be better. 
% For this reason, we define a reduction which can be applied to $\Gamma_{data}$, and which is performed as a post-transformation. This reduction, could propose better performance depending on the used network speed.

% \begin{myprop}
% Denoting four computations $c_l$, $c_k$, $k_i$ and $k_j$ of $\Gamma$, with $l<k$, $i<j$, and where $c_l \pprec c_k$ and $k_i \pprec k_j$. Then, computations $c_k$ and $k_j$ are tansformed to $c^*_k(c_k,update(w_l))$ and $c^*_j(k_j,update(w_i))$. If $i<k$, a single update call can be performed at $k$ such that $c^*_k(c_k,update(w_l\cup w_i))$, and in this case, the computation $k_j$ is not transformed.
% \end{myprop}

% \begin{proof}
% By definition, the synchronization of $w_i$ can be performed between $k_i$ and $k_j$. As $i<k$ the computation $k_i$ is not performed between $k$ and $j$. As a result, the synchronization of $w_i$ can be performed between $k_i$ and $c_k$.
% \end{proof}

% In other words, for each synchronization in $\Gamma_{data}$, if the computation $k_i$ from which $k_j$ is synchronized occurs before the last synchronization, the two synchronizations can be reduced to one with the union of the data to update.

 The final step of this parallelization is to run $\Gamma_{sync}$ on each resource. Thus, for each resource $0 \leq k \leq p-1$ a multi-stencil program, defined by
\begin{equation}
\mathcal{MSP}_k(T,\mathcal{M}_k,\mathcal{E},\mathcal{D},\Delta_k,\mathcal{S},\Gamma_{sync}),
\end{equation}
is performed.

%We denote this parallelization technique a coarse-grain data parallelization in contrast with the same technique applied to the finer level of a single computation. In this case, for a computation $c(R,w,D,\text{exp})$, the local domain $D$ is partitionned for $p$ resources $\{D_0,\dots,D_{p-1}\}$, and each resource $k$ is responsible for a sub-computation $c_k(R,w,D_k,\text{exp}')$.
%This finer data parallelization technique is not directly adressed by the work presented in this paper, but is exhibited, as it will be explained later.

@@

\begin{mydef}
For two computations $k_i(R_i,w_i,\text{exp}_i,D_i)$ and $k_j(R_j,w_j,\text{exp}_j,D_j)$, with $i < j$, it is said that $k_j$ is data dependant from $k_i$, denoted $k_i\prec k_j$, if $\{w_i\} \cap R_j \neq \emptyset$. In this case, $k_i$ has to be computed before $k_j$. The binary relation $k_i\prec k_j$ represents a \textit{dependency}.
\end{mydef}

% \begin{myprop}
% The binary relation $\prec$ is not transitive. 
% \end{myprop}

% \begin{proof}
% Considering three computations $c_0(R_0,w_0,\text{exp}_0)$, $c_1(R_1,w_1,\text{exp}_1)$ and $c_2(R_2,w_2,\text{exp}_2)$, where $\{w_0\}\cap R_1 \neq \emptyset$ and $\{w_1\}\cap R_2 \neq \emptyset$. In this case two dependencies can be extracted $c_0 \prec c_1$ and $c_1 \prec c_2$. However, the dependency $c_0 \prec c_2$ is not true as $\{w_0\}\cap R_2 = \emptyset$.
% \end{proof}

% It has been proved that the binary relation $\prec$ is not transitive. This property is due to the fact that $\prec$ is defined as a data dependency between computations. However, another type of dependency, only due to time, is created from $\prec$.

% \begin{mydef}
% For two computations $k_i(R_i,w_i,\text{exp}_i)$ and $k_j(R_j,w_j,\text{exp}_j)$, a time dependency is denoted $k_i \blacktriangleleft k_j$ and means that $k_i$ has to be computed before $k_j$.
% \end{mydef}

% As a result, the relation $\blacktriangleleft$ is more general than $\prec$.

% \begin{myprop}
% For two computations $k_i(R_i,w_i,\text{exp}_i)$ and $k_j(R_j,w_j,\text{exp}_j)$ such that $k_i \prec k_j$, $k_i \blacktriangleleft k_j$ is verified.
% \end{myprop}

% \begin{proof}
% By definition $k_i \prec k_j$ means that $\{w_i\}\cap R_j \neq \emptyset$ and that $k_i$ has to computed before $k_j$. %Thus, the relation $k_i \prec k_j$ is stronger than $k_i \blacktriangleleft k_j$.
% \end{proof}

% \begin{myprop}
% The binary relation $\blacktriangleleft$ is transitive. 
% \end{myprop}

% \begin{proof}
% Considering three computations $c_0(R_0,w_0,\text{exp}_0)$, $c_1(R_1,w_1,\text{exp}_1)$ and $c_2(R_2,w_2,\text{exp}_2)$, where $c_0 \blacktriangleleft c_1$ and $c_1 \blacktriangleleft c_2$. $c_0$ is computed before $c_1$ and $c_1$ is computed before $c_2$, as a result $c_0$ is computed before $c_2$ and the relation $c_0 \blacktriangleleft c_2$ is verified. 
% \end{proof}

@@

From an ordered list of computations $\Gamma_{sync}$, a directed dependency graph $\Gamma_{dep}(V,A)$ can be built finding all pairs of computations $k_i(R_i,w_i,\text{exp}_i,D_i)$ and $k_j(R_j,w_j,\text{exp}_j,D_j)$, with $i<j$, such that $k_i \prec k_j$. %Those time dependencies can be found from pairs of data dependencies $k_i \prec k_j$.

@@

\begin{mydef}
From the ordered list $\Gamma_{sync}$ of computations $c(R,w,\text{exp},D)$, a directed dependency graph $\Gamma_{dep}(V,A)$ is obtained from the call $T_{dep}(\Gamma_{sync},0)$, where $F_{dep}$ is the recursive function
\begin{equation*}
F_{dep}(\Gamma_{sync},j) = 
\begin{cases} 	(\{\},\{\}) & \mbox{if }j=|\Gamma_{sync}|\\
				(k_j, \{(\overset{\frown}{k_i,k_j})\mbox{, }\forall i < j \mbox{, } k_i\prec k_j \})\cup F_{dep}(\Gamma_{sync},j+1) & \mbox{if }j<|\Gamma_{sync}|
\end{cases}
\end{equation*}
\end{mydef}

@@

As a result, the hybrid parallelization of a multi-stencil program $\mathcal{MSP}$, produces $k$ multi-stencil programs
\begin{equation*}
\mathcal{MSP}_k(T,\mathcal{M}_k,\mathcal{E},\mathcal{D},\Delta_k,\Gamma_{dep}).
\end{equation*}
The set of computations $\Gamma_{dep}$ is a dependency graph between computations $k_i$ of $\Gamma$ and the synchronization computations $synk_i$ added into $\Gamma_{sync}$. $\Gamma_{dep}$ can be built from the call to 
\begin{equation*}
F_{dep}(F_{sync}(\Gamma,0),0).
\end{equation*}

% Using the function $T_{task}$ to build $\Gamma_{task}$, however, duplication of dependencies may occur because of the transitivity of the relation $\blacktriangleleft$. Actually, as the relation $\prec$ verifies the relation $\blacktriangleleft$, and as $\blacktriangleleft$ is transitive, if for three computations $c_k(R_k,w_k,\text{exp}_k)$, $k_i(R_i,w_i,\text{exp}_i)$ and $k_j(R_j,w_j,\text{exp}_j)$, with $k<i<j$ and where $c_k \prec k_i$, $k_i \prec k_j$ and $c_k \prec k_j$, then the transitivity $c_k \blacktriangleleft k_j$ is verified because of the relations $c_k \blacktriangleleft k_i$ and $k_i \blacktriangleleft k_j$. However, $c_k \blacktriangleleft k_j$ is also directly represented by the binary relation $c_k \prec k_j$. As a result, a duplication of the dependency is created in $\Gamma_{task}$. In Figure~\ref{fig:duplication} an example of duplication is given for the relation $c_1 \blacktriangleleft c_4$.

% \begin{figure}[h!]
% \begin{center}
% \begin{tikzpicture}[shorten >=1pt, node distance=2cm, on grid, auto]
%    \node[] (c0) at (0,0) {$c_0$};
%    \node[] (c1) at (1,0) {$c_1$};
%    \node[] (c2) at (2,0) {$c_2$};
%    \node[] (c3) at (3,0) {$c_3$};
%    \node[] (c4) at (4,0) {$c_4$};

%   \path[->]
%     (c0) edge node {} (c1)
%     (c1.east) edge node {} (c2)
%     	 edge [bend right=50] node [swap] {} (c4)
%     (c2) edge node {} (c3)
%     (c3) edge node {} (c4);
%   \end{tikzpicture}
%   \caption{Useless duplication of the dependency $c_1 \blacktriangleleft c_4$}
%   \label{fig:duplication}
% \end{center}
% \end{figure}

% Another view of the relations $\prec$ and $\blacktriangleleft$ is that $k_i\blacktriangleleft k_j$ in $\Gamma_{task}$ is a path from $k_i$ to $k_j$ of any size, while $k_i \prec k_j$ is a path of size $1$ from $k_i$ to $k_j$. And in any case a path in the graph represents a time dependency between computations. Scheduling a graph dependency which do not contains duplication of information is easier. 
% %For this reason, we define a reduction to apply to $\Gamma_{task}$, and which is performed as a post-transformation. In the example of Figure~\ref{fig:duplication} the dependency represented by the path $c_1 \blacktriangleleft c_2 \blacktriangleleft c_3 \blacktriangleleft c_4$ gives more information than the one directly represented by $c_1 \prec c_4$. The proposed reduction is based on this property.

% % \begin{mydef}
% % The dependency graph $\Gamma_{task}(V,A)$ is reduced to the dependency graph $\Gamma_{task}(V,A')$, where $A' \subseteq A$. A direct arc $(\overset{\frown}{k_i,k_j})$, $i<j$, of $A$, which means that $k_i \prec k_j$, is included in $A'$ if no longer path $k_i \blacktriangleleft k_j$ exists in $A$ because of transitivity.
% % \end{mydef}
% \begin{mydef}
% A DAG is transitive if it contains an arc $(\overset{\frown}{u,v})$ between any two vertices such that there is a path from $u$ to $v$. The transitive closure of a DAG $G=(V,E)$, is the DAG $G_T=(V,E_T)$ for which $E_T$ is the minimal subset of $V \times V$ that includes $E$ and makes $G_T$ transitive.
% \end{mydef}

% \begin{mydef}
% An arc $(\overset{\frown}{u,v})$ of a DAG is redundant if there is a path from $u$ to $v$ in the DAG that does not include the edge. A DAG that does not contain any redundant arc is called minimal. The transitive reduction of a DAG $G$ is its unique minimal sub-graph having the same transitive closure.
% \end{mydef}

% Thus, the removal of the redundant dependencies of $\Gamma_{task}$ consists in applying a transitive reduction.

%------------------------------
% \subsection{Hybrid parallelism}
% It is also possible to combine data and task parallelization techniques to get hybrid parallelism, sometimes more efficient on hybrid architectures. A coarse-grain data parallelism creates for $k$ resources $k$ multi-stencil computations $\mathcal{MSP}_k(T,\mathcal{M}_k,\Delta_k,\Gamma_{data})$, where $\Gamma_{data}$ is built from an ordered list of computations and is itself an ordered list of computations $k_i$ and updated-computations $*_i$. On the other hand, the task parallelization technique builds, from an ordered list of computation, a graph dependency $\Gamma_{task}$. As a result $T_{data}$ and $T_{task}$ can be composed to build $\Gamma_{hybrid}$.

% However, we have to clarify the dependency operator $\prec$ for the specific case of updated-computations. As explained in Section~\ref{sect:dataparal}, if $k_j$ is an updated-computations, which means that at least one $k_i$ exists such that $i<j$ and $k_i \pprec k_j$, $k_j$ is transformed to the ordered list of two computations

% \begin{equation*}
% [*_j(R^*_j,w^*_j),c^*_j(R_j \cup \{w^*_j\},w_j,\text{exp}_j)].
% \end{equation*}

% As a result, and because $*_j$ is itself a computation, it appears dependencies relations. First, by definition, the dependency $c^*_j \prec *_j$ is created because $\{w^*_j\} \subset R_j \cup \{w^*_j\}$. Second, as $R^*_j = \{\bigcup_{i} \{w_i\} \text{, }\forall i<j\text{, and }k_i \pprec k_j\}$, for each $i<j$ the dependency $k_i \prec *_j$ is also created.

% The choice of separating $*_j$ from $k_j$ has been done to entirely expose the eventual task parallelism in the hybrid parallelization. For example, Figure~\ref{fig:more} give an exemple where initially $c_0 \pprec c_2$ and $c_0 \prec c_1$, $c_1 \prec c_2$. Thus, $c_2$ is replaced first in $\Gamma_{data}$ by $[*_2(\{w_0\},w^*_2),c^*_2(R_2 \cup \{w^*_2\},\text{exp}_2)]$. As a result, when applying $T_{task}$ on $\Gamma_{data}$, it appears that $*_2$ can be performed in parallel with $c_1$. This is due to the fact that $*_2$ is splitted from from $c_2$ and that $R^*_2=\{w_0\}$ is a subset of $R_2=\{w_0,w_1\}$ and not equal to $R_2$.

% \begin{figure}[h!]
% \begin{center}
% \begin{tikzpicture}[shorten >=1pt, node distance=2cm, on grid, auto]
%    \node[] (c0) at (0,0) {$c_0$};
%    \node[] (c1) at (2,0) {$c_1$};
%    \node[] (star2) at (2,1) {$*_2$};
%    \node[] (c2) at (4,0) {$c^*_2$};

%   \path[->]
%     (c0) edge node {} (c1)
%          edge node {} (star2)
%     (c1) edge node {} (c2)
%     (star2) edge node {} (c2);
%   \end{tikzpicture}
%   \caption{Example of hybrid parallelization.}
%   \label{fig:more}
% \end{center}
% \end{figure}

% As a result, the hybrid parallelization of a multi-stencil program $\mathcal{MSP}$, produces $k$ multi-stencil programs
% \begin{equation*}
% \mathcal{MSP}_k(T,\mathcal{M}_k,\Delta_k,\Gamma_{hybrid}).
% \end{equation*}
% The set of computations $\Gamma_{hybrid}$ is a dependency graph between computations $k_i$ and update computations $*_i$, and can be built from the call to 
% \begin{equation*}
% T_{task}(T_{data}(\Gamma,0),0).
% \end{equation*}

% The rest of this paper deals with the hybrid parallelization of multi-stencil programs described above. The reason of this choice is the fact that by adressing the hybrid parallelism case, we also are able to adress data parallelism and task parallelism, from which hybrid parallelism is built.

>>>>

\begin{mydef}
A synchronization between two computations $k_i \pprec k_j$ is defined as a specific computation 
\begin{equation*}
k_{sync}(S,R,w,comp)
\end{equation*}
\end{mydef}
where $S=\emptyset$, $R=\{(w_i,n_j)\}$, $w=(w_i,n_j)$ and $comp=identity$. In other words, $w_i$ has to be synchronized for the neighborhood $n_j$.

\begin{mydef}
If $k_i \pprec k_j$, $k_j$ is replaced by the list
\begin{equation*}
[k_{sync}(\emptyset,\{(w_i,n_j)\},(w_i,n_j),identity), k_j]
\end{equation*}
\end{mydef}

@@

\begin{mydef}
For a reduction kernel $k(S,R,(w,D),comp)$, if we denote by $w^i$, $0 \leq i<p$, the local $w$ computed on the resource $i$, a reduction synchronization is defined as the specific computation 
\begin{equation*}
k_{sync}(S,R_s,w,comp)
\end{equation*}
\end{mydef}
where, $S=\emptyset$, $R=\{(w^0,entity(w^0)) \dots (w^{p-1},entity(w^{p-1}))\}$ and $w$ and $comp$ are the same than the actual reduction.

@@

\begin{mydef}
From the ordered list of computation $\Gamma$, a new ordered list $\Gamma_{sync}$ is obtained from the call $\Gamma_{sync} = F_{sync}(\Gamma,0)$, where $F_{sync}$ is the recursive function defined as
\begin{equation*}
F_{sync}(\Gamma,j) = 
\begin{cases} 	[] \mbox{ if }j=|\Gamma|\\
				[k_{sync}(\emptyset,\{w_i\},w_i,identity,\emptyset), k_j] \cdot F_{sync}(\Gamma,j+1)\\
				\text{ } \qquad \forall i<j \text{, and }k_i \pprec k_j\\
				[k_j, k_{sync}(\emptyset,,w_j,comp_j)] \cdot F_{sync}(\Gamma,j+1)\\ 
				\text{ } \qquad \mbox{if } w_j \in \mathcal{S} \text{, and } R_j=\{(r_j,entity(r_j))\}\\
				[\Gamma[j]] \cdot F_{sync}(\Gamma,j+1) \text{ else}.
\end{cases}
\end{equation*}
\end{mydef}

 The final step of this parallelization is to run $\Gamma_{sync}$ on each resource. Thus, for each resource $0 \leq m \leq p-1$ a multi-stencil program, defined by
\begin{equation}
\mathcal{MSP}_m(T,\mathcal{M}_m,\mathcal{E}_m,\mathcal{D}_m,\Delta_m,\mathcal{S},\Gamma_{sync}),
\end{equation}
is performed.

@@

\begin{mydef}
For two computations $k_i(S_i,R_i,(w_i,D_i),comp_i)$ and $k_j(S_j,R_j,(w_j,D_j),comp_j)$, with $i < j$, it is said that $k_j$ is data dependant from $k_i$, denoted $k_i\prec k_j$, if $w_i \in R_j$. In this case, $k_i$ has to be computed before $k_j$. The binary relation $k_i\prec k_j$ represents a \textit{dependency}.
\end{mydef}

@@

From an ordered list of computations $\Gamma_{sync}$, a directed dependency graph $\Gamma_{dep}(V,A)$ can be built finding all pairs of computations $k_i(S_i,R_i,(w_i,D_i),comp_i)$ and $k_j(S_j,R_j,(w_j,D_j),comp_j)$, with $i<j$, such that $k_i \prec k_j$.

@@

\begin{mydef}
From the ordered list $\Gamma_{sync}$ of computation kernels, a directed dependency graph $\Gamma_{dep}(V,A)$ is obtained from the call $T_{dep}(\Gamma_{sync},0)$, where $F_{dep}$ is the recursive function
\begin{equation*}
F_{dep}(\Gamma_{sync},j) = 
\begin{cases} 	(\{\},\{\}) \mbox{ if }j=|\Gamma_{sync}|\\
				(k_j, \{(\overset{\frown}{k_i,k_j})\mbox{, }\forall i < j \mbox{, } k_i\prec k_j \})\\
				\text{ } \qquad \cup F_{dep}(\Gamma_{sync},j+1) \mbox{ if }j<|\Gamma_{sync}|
\end{cases}
\end{equation*}
\end{mydef}

@@

As a result of the hybrid parallelization, each resource $0 \leq m \leq p-1$ perform a multi-stencil program, defined by
\begin{equation*}
\mathcal{MSP}_m(T,\mathcal{M}_m,\mathcal{E}_m,\mathcal{D}_m,\Delta_m,\Gamma_{dep}).
\end{equation*}
The set of computations $\Gamma_{dep}$ is a dependency graph between computations $k_i$ of $\Gamma$ and the synchronization kernels added into $\Gamma_{sync}$. $\Gamma_{dep}$ can be built from the call to 
\begin{equation*}
F_{dep}(F_{sync}(\Gamma,0),0).
\end{equation*}
